---
title: "Draft2"
output: slidy_presentation
date: '2022-04-18'
---

# Load libraries 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(plotly)
```

# Add in Data 
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
deaths_df

colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", 
                         "indoor_deaths", "outdoor_deaths", "ozone_deaths")

glimpse(deaths_df)
```

# Questions to Consider 

## Which country has the greatest number of deaths? The least? (dependent on population)

```{r, include=FALSE}
# Want the range 2007-2017 
#Maybe try increments...?

# Look at the trend of deaths within the last decade from low and high populated countries
death_07_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2007, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Populated_Deaths_07 = mean(total_deaths))
death_07_high

death_07_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2007, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Populated_Deaths_07 = mean(total_deaths))
death_07_low

death_12_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2012, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Populated_Deaths_12 = mean(total_deaths))
death_12_high

death_12_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2012, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Populated_Deaths_12 = mean(total_deaths))
death_12_low

death_17_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Populated_Deaths_17 = mean(total_deaths))
death_17_high

death_17_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(Low_Populated_Deaths_17 = mean(total_deaths))
death_17_low

#cohesive table so it's easier to visualize 
#Low Population Table 2007-2017
all_low <- death_07_low %>% 
  mutate(death_12_low) %>% 
  mutate(death_17_low)
all_low

#High Population Table 2007-2017
all_high <- death_07_high %>% 
  mutate(death_12_high) %>% 
  mutate(death_17_high)
all_high

#Then graph it...but how?? 
#Scatter plot (color by country, info pop up like the average death and by year)

```
# Separation by Population
```{r, echo=FALSE}
#SCRATCH WORK ABOVE 

#Not sure if we want to use the selected few countries we looked at earlier?
#Or if we wanted to use this as a overview of the data
library(knitr)
#install.packages('kableExtra')
library(kableExtra)

#separation by high vs low population among continents 
death_07_high <- deaths_df %>% 
  filter(year == 2007, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_07_high = mean(total_deaths))
#death_07_high


death_07_low<- deaths_df %>% 
  filter(year == 2007, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_07_low = mean(total_deaths))
#death_07_low

#Side by side table 
#kable(list(death_07_high, death_07_low))
kable(death_07_high) %>% 
  kable_styling(full_width = FALSE, position = 'float_left')
kable(death_07_low) %>% 
  kable_styling(full_width = FALSE, position = 'left')

#dataframe for 2007 high vs low - pop countries
df_07<- data.frame(death_07_high, death_07_low)
df_07
colnames(df_07)
colnames(df_07) <- c("high_pop_country", "average_death_07_high", 
                     "low_pop_country", "average_death_07_low")
df_07


death_12_high <- deaths_df %>% 
  filter(year == 2012, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_12_high = mean(total_deaths))
#death_12_high

death_12_low<- deaths_df %>% 
  filter(year == 2012, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_12_low = mean(total_deaths))
#death_12_low

#Combo table 2012
#kable(list(death_12_high, death_12_low))
kable(death_12_high) %>% 
  kable_styling(full_width = FALSE, position = 'float_left')
kable(death_12_low) %>% 
  kable_styling(full_width = FALSE, position = 'left')

#Data frame for 2012
df_12<- data.frame(death_12_high, death_12_low)
df_12

colnames(df_12) <- c("high_pop_country", "average_death_12_high", 
                     "low_pop_country", "average_death_12_low")
df_12

death_17_high <- deaths_df %>% 
  filter(year == 2017, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_17_high = mean(total_deaths))
#death_17_high

death_17_low<- deaths_df %>% 
  filter(year == 2017, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_17_low = mean(total_deaths))
#death_17_low

#Combo table 2017
#kable(list(death_17_high, death_17_low))
kable(death_17_high) %>% 
  kable_styling(full_width = FALSE, position = 'float_left')
kable(death_17_low) %>% 
  kable_styling(full_width = FALSE, position = 'left')

#Data frame for 2017
df_17<- data.frame(death_17_high, death_17_low)
df_17

colnames(df_17) <- c("high_pop_country", "average_death_17_high", 
                     "low_pop_country", "average_death_17_low")
df_17

#Graph selected countries 
#Is there a way to color by low and high and combine the two graphs below?
# HEAD
#I removed combo_07 so this might not work... 

#combo_07 %>% 
  #ggplot(aes(x = country, y = average_death_07_low)) + 
  #geom_point()

#combo_07 %>% 
  #ggplot(aes(x = country, y = average_death_07_high)) + 
  #geom_point()
# 
#combo_07 %>% 
#  ggplot() + 
#  geom_col(mapping = aes(x = country, y = average_death_07_low, fill = country))+
#             xlab("Country")+
#             ylab("Average deaths (per 100,000)")+
#             ggtitle("Average total deaths in low-population countries")+
#  coord_flip()

#combo_07 %>% 
#  ggplot() + 
#  geom_col(mapping = aes(x = country, y = average_death_07_high, fill = country))+
#             xlab("Country")+
#             ylab("Average deaths (per 100,000)")+
#             ggtitle("Average total deaths in high-population countries")+
#  coord_flip()

ggplot(df_07)+
  geom_col(mapping = aes(x=high_pop_country, y=average_death_07_high))+
             xlab("Country")+
             ylab("Average deaths (per 100,000)")+
             ggtitle("Average total deaths in high-population countries")+
  coord_flip()

ggplot(df_07)+
  geom_col(mapping = aes(x=low_pop_country, y=average_death_07_low))+
             xlab("Country")+
             ylab("Average deaths (per 100,000)")+
             ggtitle("Average total deaths in low-population countries")+
  coord_flip()
  
#f55e1c84fceb2d108c13d56423a26347f3c769e6
```

## Which type of air pollution has the greatest number of deaths (indoor, outdoor, particulate, or outdoor ozone)?

```{r}
glimpse(deaths_df)

#This is the average of all countries by type of air pollutant death
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))

#Graphs of each type of air pollutant:
#Will work on making it more clear, but I think the distribution is cool
#Indoor 
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths),
            avg_ozone = mean(ozone_deaths)) %>% 
  ggplot()+
  geom_point(mapping = aes(x=avg_indoor, y=country))+
  ggtitle("Indoor Air Pollutant Deaths")+
  xlab("Deaths per 100,000")+
  ylab("Country")

#Outdoor
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths),
            avg_ozone = mean(ozone_deaths)) %>% 
  ggplot()+
  geom_point(mapping = aes(x=avg_outdoor, y=country))+
  ggtitle("Outdoor Air Pollutant Deaths")+
  xlab("Deaths per 100,000")+
  ylab("Country")

#Ozone
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths),
            avg_ozone = mean(ozone_deaths)) %>% 
  ggplot()+
  geom_point(mapping = aes(x=avg_ozone, y=country))+
  ggtitle("Ozone Air Pollutant Deaths")+
  xlab("Deaths per 100,000")+
  ylab("Country")

#Having fun with plots
#total deaths
deaths_df %>% 
    filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 
                          'Israel', 'Tonga','United States', 'Brazil', 
                          'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country, year, total_deaths) %>% 
  ggplot()+
  geom_point(aes(year, total_deaths, color = country))+
  ggtitle("Total deaths throughout the years in sample countries")+
  xlab("Year")+
  ylab("Deaths per 100,000")


#indoor deaths
deaths_df %>% 
    filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 
                          'Israel', 'Tonga','United States', 'Brazil', 
                          'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Indoor deaths throughout the years in sample countries")+
  xlab("Year")+
  ylab("Deaths per 100,000")

#ozone deaths
p2 <- deaths_df %>% 
    filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 
                          'Israel', 'Tonga','United States', 'Brazil', 
                          'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_line(aes(year, ozone_deaths, color = country))+
  ggtitle("Ozone deaths throughout the years in sample countries")+
  scale_y_log10() + 
  xlab("Year")+
  ylab("Deaths per 100,000") 
ggplotly(p2)
  
  
#Sort descending to see which country had the highest of each...(not sure if this is right)
#does separating the arrange individually fix it??
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
  arrange(desc(avg_indoor), desc(avg_outdoor), desc(avg_ozone))

library(plotly)
p <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
ggplotly(p)
```

## Which year had the worst air pollution (average the three diff types)?

```{r}
deaths_df %>% 
  group_by(country, year) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
  arrange(desc(avg_indoor)) %>% 
  arrange(desc(avg_outdoor)) %>% 
  arrange(desc(avg_ozone))
```

## Which year had the worst indoor? Outdoor particulate? Outdoor ozone?

<<<<<<< HEAD
##Focus in on some countries/ continents

##Pull in more data on populations for the selected countries

##Industrial and non industrial zones, what countries have coal plants? oil plants? 
```{r}

```


##Years to focus on: 1990 to 2017, 

#start with high population country and to choose the low population, choose a country that is 10% of that high population and choose a country closest (2020 population)

##Countries:
#US - north america - high pop
#Canada - north america - low pop

#Brazil - S america - high pop
#Chile - S America - low pop

#Nigeria - Africa - high pop
#Malawi - Africa - low pop

#Germany - Europe - high pop
#Serbia - Europe - low pop

#China - Pakistan - high pop 
#Israel - Sri Lanka - low pop

#Australis - Oceania - high pop
#New Zealand - Oceania - low pop
=======
```{r}
#Probably need to fix the above to answer this one 


```

# Notes from first meeting 

## Hone in to only one year. Find data that is only population, continent data. Choose a couple countries/continents to work on 

## Look at industrialized vs non-industrialized countries 
#>>>>>>> a099fd8818f069a204f274675f043d2f94067c4a



#Alison's scratch work! :) 


#Filtering to desired countries
```{r}
#north america
america <- deaths_df %>% filter(country == "United States" & year > 1999)
canada <- deaths_df %>% filter(country == "Canada" & year > 1999)


#south america 
brazil <- deaths_df %>% filter(country == "Brazil" & year > 1999)
argentina <- deaths_df %>% filter(country == "Argentina" & year > 1999)

#Africa
ethiopia <- deaths_df %>% filter(country == "Ethiopia" & year > 1999)
seychelles <- deaths_df %>% filter(country == "Seychelles" & year > 1999)

#Europe
russia <- deaths_df %>% filter(country == "Russia" & year > 1999)
romania <- deaths_df %>% filter(country == "Romania" & year > 1999)

#asia
china <- deaths_df %>% filter(country == "China" & year > 1999)
israel <- deaths_df %>% filter(country == "Israel" & year > 1999)

#Oceania
australia <- deaths_df %>% filter(country == "Australia" & year > 1999)
tonga <- deaths_df %>% filter(country == "Tonga" & year > 1999)

```

#
```{r}



```


