---
title: "Draft2"
output: slidy_presentation
date: '2022-04-18'
---

# Load libraries 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# Add in Data 
```{r}
read_csv('death-rates-from-air-pollution.csv')
```

# Questions to Consider 

## Which country has the greatest number of deaths? The least?

## Which type of air pollution has the greatest number of deaths (indoor, outdoor, particulate, or outdoor ozone)?

## Which year had the worst air pollution (average the three diff types)?

## Which year had the worst indoor? Outdoor particulate? Outdoor ozone?

##Focus in on some countries/ continents

##Pull in more data on populations for the selected countries

##Industrial and non industrial zones, what countries have coal plants? oil plants? 


##Years to focus on: 2007 to 2017, 

##Countries:
#US - north america - high pop
#Canada - north america - low pop

#Brazil - S america - high pop
#Argentina - S America - low pop

#Ethiopia - Africa - high pop
#Seychellas - Africa - low pop

#Russia - Europe - high pop
#Romania - Europe - low pop

#China - Asia - high pop 
#Israel - Asia - low pop

#Australis - Oceania - high pop
#Tonga - Oceania - low pop
