---
title: "Draft2"
output: slidy_presentation
date: '2022-04-18'
---

# Load libraries 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# Add in Data 
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
deaths_df

colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", 
                         "indoor_deaths", "outdoor_deaths", "ozone_deaths")

glimpse(deaths_df)
```

# Questions to Consider 

## Which country has the greatest number of deaths? The least? (dependent on population)

```{r}
# Want the range 2007-2017 
# took this out c(2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017)

#Maybe try increments...?
#When looking at it you can see a decrease among the three years 
#death_07<- deaths_df %>% 
  #group_by(country) %>% 
  #filter(year == 2007) %>% 
  #select(total_deaths) %>% 
  #summarize(average_death_07 = mean(total_deaths))

#death_12<- deaths_df %>% 
  #group_by(country) %>% 
  #filter(year == 2012) %>% 
  #select(total_deaths) %>% 
  #summarize(average_death_12 = mean(total_deaths))

#death_17<- deaths_df %>% 
  #group_by(country) %>% 
  #filter(year == 2017) %>% 
  #select(total_deaths) %>% 
  #summarize(average_death_17 = mean(total_deaths))

#try to have one cohesive table so it's easier to visualize 
#combo <- death_07 %>% 
  #mutate(death_12) %>% 
  #mutate(death_17)

#combo

#Then graph it 
#Just realized there's 200+ countries....the x-axis looks terrible 
#Would be nice to color by continent but then that would mean going through each country
#ggplot(death_07, aes(x= country, y=average_death_07)) +
  #geom_point() +
  #theme(axis.text.x = element_text(angle = 15)) #yah..no rotation made it worse


```

```{r}
#SCRATCH WORK ABOVE 

#Not sure if we want to use the selected few countries we looked at earlier?
#Or if we wanted to use this as a overview of the data

#separation by high vs low population among continents 
death_07_high <- deaths_df %>% 
  filter(year == 2007, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_07_high = mean(total_deaths))


death_07_low<- deaths_df %>% 
  filter(year == 2007, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_07_low = mean(total_deaths))

#Combo table 2007
combo_07<- death_07_high %>% 
  mutate(death_07_low)
combo_07

death_12_high <- deaths_df %>% 
  filter(year == 2012, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_12_high = mean(total_deaths))


death_12_low<- deaths_df %>% 
  filter(year == 2012, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_12_low = mean(total_deaths))

#Combo table 2012
combo_12<- death_12_high %>% 
  mutate(death_12_low)
combo_12

death_17_high <- deaths_df %>% 
  filter(year == 2017, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_17_high = mean(total_deaths))


death_17_low<- deaths_df %>% 
  filter(year == 2017, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(average_death_17_low = mean(total_deaths))


#Combo table 2017
combo_17<- death_17_high %>% 
  mutate(death_17_low)
combo_17

#Graph selected countries 
#Is there a way to color by low and high and combine the two graphs below?
combo_07 %>% 
  ggplot(aes(x = country, y = average_death_07_low)) + 
  geom_point()

combo_07 %>% 
  ggplot(aes(x = country, y = average_death_07_high)) + 
  geom_point()
```

## Which type of air pollution has the greatest number of deaths (indoor, outdoor, particulate, or outdoor ozone)?

```{r}
glimpse(deaths_df)

#This is the average of all countries by type of air pollutant death
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
  
#Sort descending to see which country had the highest of each...(not sure if this is right)
#does separating the arrange individually fix it??
deaths_df %>% 
  group_by(country) %>% 
  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
  arrange(desc(avg_indoor), desc(avg_outdoor), desc(avg_ozone))
```

## Which year had the worst air pollution (average the three diff types)?

```{r}
deaths_df %>% 
  group_by(country, year) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
  arrange(desc(avg_indoor)) %>% 
  arrange(desc(avg_outdoor)) %>% 
  arrange(desc(avg_ozone))
````

## Which year had the worst indoor? Outdoor particulate? Outdoor ozone?

```{r}
#Probably need to fix the above to answer this one 


````

# Notes from first meeting 

## Hone in to only one year. Find data that is only population, continent data. Choose a couple countries/continents to work on 

## Look at industrialized vs non-industrialized countries 
