---
title: "Final Report"
author: "Alison Cheek & Elizabeth Bekele"
date: '2022-05-06'
output: html_document
---
We were excited to do our report over this data because it was relatively tidy and had quite a few categorical variables and options for additional columns to graph.


# Introduction 
* Air pollution can be detrimental to both our health and the climate
  + Outdoor and indoor air pollution cause chronic pain, respiratory diseases, shortened lifespan
  + Air pollution kills about 7 million people worldwide every year 
  + Hopefully this information can showcase the importance of air pollution and that we should be more mindful about our planet
  
* Overview
  + We will see how different air pollution types affect the population 
  + compare past and present population numbers
  + determine which air pollutant type has the highest associated death rate


# Packages Required 
```{r, message=FALSE, warning=FALSE}
#This will allow us to filter through our data 
library(tidyverse)
library(dplyr)
#This will help us plot figures to showcase our findings
library(ggplot2)
#This will help us organize and display our data as necessary 
library(knitr)
library(kableExtra)
#This expands our plot uses 
library(plotly)
#Scientific Notation Disabled 
options(scipen=T)
```


# Deaths Data
* Deaths from pollution:
  + Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
  + Author: Akshat Giri
  + Last update: 2 years ago
  
Import the deaths-due-to-air-pollution data
```{r}
deaths_df_old <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
glimpse(deaths_df_old)
```
Fixed: use rename instead of colnames

We are going to rename a few of the columns and glimpse the data
```{r}
deaths_df<- deaths_df_old %>% rename(country=Entity, acronym=Code, year=Year, total_deaths=Air.pollution..total...deaths.per.100.000., indoor_deaths=Indoor.air.pollution..deaths.per.100.000., outdoor_deaths=Outdoor.particulate.matter..deaths.per.100.000., ozone_deaths=Outdoor.ozone.pollution..deaths.per.100.000.)

glimpse(deaths_df)
```

# Data Variables
Variables that interest us here include: 

* __country__ 
* __total_deaths__: per 100,000
* __indoor_deaths__: Indoor air pollution is considered pollution that occurs in the household. Cooking with solid fuels:
  + Wood
  + Crop waste, dung
  + Charcoal, coal
* __outdoor_deaths__: Outdoor air or ambient air are emissions caused by combustion processes from motor vehicles, solid fuel burning and industry 
  + Ozone (O3)
  + Particulate matter (PM10 and PM2.5)
  + Nitrogen dioxide (NO2)
  + Carbon monoxide (CO)
  + Sulfur dioxide (SO2)
* __ozone_deaths__: Ozone is a gas that occurs both in Earth's upper atmosphere and at ground level. Ozone in the atmosphere is an important and helpful greenhouse gas, but ground-level ozone is created by extensive use of fossil fuels: 
  + Pollutants emitted by cars
  + Power plants, industrial boilers, refineries, chemical plants

# World Population Data 
* World Population:
  + Kaggle: [https://www.kaggle.com/datasets/imdevskp/world-population-19602018] 
  + Author: Devakumar K. P. 
  + Last update: 2 Years ago


Now, let's take a look at the population data.
```{r}
world_pop <- read.csv("population_total_long.csv")
glimpse(world_pop)
```

***

To get a general idea of 'deaths-dataframe' we made, let's make a plots to see what's happening. This is a plot of indoor x outdoor deaths around the world by country.

```{r, echo=FALSE}
d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point() +
  ggtitle("Outdoor Deaths vs Indoor Deaths")
ggplotly(d)
```

This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.

***
We selected a high population from each continent and used the formula below to determine the low population.

Low population = high population * .10
  
  - We wanted to have a consistent variation between our selection of low and high populated countries. 
  - Based on our calculation we selected the low populated country that was closest to 10% of the high populated country.
  - EX: U.S (high populated) = 331002651. If we multiply by .10 we get 33100265.1 and Canada (low populated) was a population of 37742154 which was closest to 10% of the US population in 2020
  
```{r, echo=FALSE}
#selecting high-population countries from the world population data frame 
high_pop_countries <- world_pop %>% 
  filter(Year> 1995 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(Year)
  
high<- ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
ggplotly(high)

#selecting low-population countries from the world population data frame 
low_pop_countries <- world_pop %>% 
  filter(Year> 1995 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
  group_by(Year)

low<- ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
ggplotly(low)


kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))


```


# Combine Data Sets

First let's look at a table of the high and low populated countries using the world population data set.

```{r, echo=FALSE}
kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))
```
***

Next, we are going to see the death count for high and low populated countries using the deaths dataframe.

```{r, echo=FALSE}
#selecting high-population deaths from death dataframe 
high_pop_death <- deaths_df %>% 
  filter(year > 1995 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(year)


#selecting low-population deaths from death dataframe 
low_pop_death <- deaths_df %>% 
  filter(year > 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  group_by(year)

kable(head(list(high_pop_death[1:22,], low_pop_death[1:22,])))
```
***

Lastly, we will join the population and and deaths with its respected country.

```{r, message=FALSE, echo=FALSE}
#Combined High Population with Pollution Death
joined_high <- right_join(high_pop_death, high_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))


#Combined Low Population with Pollution Death
joined_low <-right_join(low_pop_death, low_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))

joined_select <- full_join(joined_high, joined_low)

kable(head(list(joined_high[1:22,], joined_low[1:22,])))
```

***

Combine the data based on continent.

```{r}
joined_all <- right_join(deaths_df, world_pop, by=c('country' = 'Country.Name', 'year' = 'Year'))
head(joined_all)

north_america <- joined_all %>% filter(country %in% c("United States", "Canada"))
head(na.omit(north_america))

south_america <- joined_all %>% filter(country %in% c("Brazil", "Chile"))
head(na.omit(south_america))

africa <- joined_all %>% filter(country %in% c("Nigeria", "Malawi"))
head(na.omit(africa))

europe <- joined_all %>% filter(country %in% c("Germany", "Serbia"))
head(na.omit(europe))

asia <- joined_all %>% filter(country %in% c("Pakistan", "Sri Lanka"))
head(na.omit(asia))

oceania <- joined_all %>% filter(country %in% c("Australia", "New Zealand"))
head(na.omit(oceania))

```


***

This is a closer view on the population growth over time in both the high and low populated countries that we selected.

```{r, echo=FALSE}
pop_high_growth<- ggplot(data=joined_high, aes(x=year, y=Count, color=country)) +
  geom_line() +
  labs(title="Population Over Time in High Populated Countries")+
  xlab("Year")+
  ylab("Population")
ggplotly(pop_high_growth)

pop_low_growth<- ggplot(data=joined_low, aes(x=year, y=Count, color=country)) +
  geom_line() +
  labs(title="Population Over Time in Low Populated Countries")+
  xlab("Year")+
  ylab("Population")
ggplotly(pop_low_growth)
```




```{r, echo=FALSE}
ggplot(data=joined_high, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
  geom_line() +
  labs(title="Percentage of Death Count in High Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
  xlab("Year")+
  ylab("Population")

ggplot(data=joined_low, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
  geom_line() +
  labs(title="Percentage of Death Count in Low Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
  xlab("Year")+
  ylab("Population")
```

# Death Count
> Which country has the highest average death count?

Let's make a table depicting the high and low populated countries and their respected death count due to pollution.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Mean total deaths of high-population countries
deaths_highpop_countries <- deaths_df %>% 
  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(hp_average_death = mean(total_deaths))

#Mean total deaths of high-population countries
deaths_lowpop_countries<- deaths_df %>% 
  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  group_by(country) %>% 
  select(total_deaths) %>% 
  summarize(lp_average_death = mean(total_deaths))

#Side-by-side Table to view deaths 
kable(list(deaths_highpop_countries, deaths_lowpop_countries))

```

> Let's see how this is different from continent to continent 

```{r}
#Mean total deaths for each continent
deaths_north <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  summarize(north_america_deaths = mean(total_deaths))


deaths_south <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  summarize(south_america_deaths = mean(total_deaths))


deaths_africa <- na.omit(africa)  %>% 
  group_by(country) %>% 
  summarize(africa_deaths = mean(total_deaths))


deaths_europe <- na.omit(europe)  %>% 
  group_by(country) %>% 
  summarize(europe_deaths = mean(total_deaths))


deaths_asia <- na.omit(asia)  %>% 
  group_by(country) %>% 
  summarize(asia_deaths = mean(total_deaths))


deaths_oceania <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  summarize(oceania_deaths = mean(total_deaths))


#Table to view continent deaths 
kable(deaths_north, caption = "North America Average Death Count")
kable(deaths_south, caption = "South America Average Death Count")
kable(deaths_africa, caption = "Africa Average Death Count")
kable(deaths_asia, caption = "Asia Average Death Count")
kable(deaths_europe, caption = "Europe Average Death Count")
kable(deaths_oceania, caption = "Oceania Average Death Count")

```



***

Here's a graph to clearly visualize the previous table

```{r, echo=FALSE, message=FALSE}
#Plot High Population Deaths (average)
h<- ggplot(deaths_highpop_countries)+
  geom_col(mapping = aes(x=country, y=hp_average_death, fill=country))+
             xlab("Country")+
             ylab("Average deaths (per 100,000)")+
             ggtitle("Average deaths in high-population countries")+
  coord_flip()

ggplotly(h)

#Plot of Low Population Deaths (average)
l <- ggplot(deaths_lowpop_countries)+
  geom_col(mapping = aes(x=country, y=lp_average_death, fill=country))+
             xlab("Country")+
             ylab("Average deaths (per 100,000)")+
             ggtitle("Average deaths in low-population countries")+
  coord_flip()

ggplotly(l)
```

*** 
So we've looked at the deaths due to pollution, but what percentage of the population was affected?

In order to get rid of the leading zeros, and clean up the y-axis, we multiplied the 'percent_high' and 'percent_low' by 100,000 since the data was per 100,000 when calculating deaths.

```{r, echo=FALSE, message=FALSE}
hp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1995) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

lp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1995) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

#Population Average Table
kable(list(hp_countries_population, lp_countries_population))

#Percentage of those affected 
affected_high<- joined_high %>% 
  mutate(percent_high = (total_deaths/mean(Count))*100000)

percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
  geom_point() +
  labs(title = "High Population Countries Affected by Air Pollution") +
  xlab("Year")+
  ylab("Percent (total_deaths/average pop)")
ggplotly(percent_high)

affected_low <- joined_low %>% 
  mutate(percent_low = (total_deaths/mean(Count))*100000)

percent_low <- affected_low %>% ggplot(aes(x = year, y = percent_low, color = country))+
  geom_point() +
  labs(title = "Low Populated Countries Affected by Air Pollution") +
  xlab("Year")+
  ylab("Percent (total_deaths/average pop)")
ggplotly(percent_low)
```



# Pollution Types
> Which type of pollution has the greatest number of deaths?

```{r, echo=FALSE, message=FALSE}
#High Population Pollutant Averages
high_poll <- deaths_df %>% 
  group_by(country) %>% 
  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
  arrange(desc(avg_indoor),desc(avg_outdoor),desc(avg_ozone))

kable(high_poll)

#Low Population Pollutant Averages 
low_poll <- deaths_df %>% 
  group_by(country) %>% 
  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))

kable(low_poll)
```


***
```{r, echo=FALSE, warning=FALSE}

#Plot High Population
# High Indoor Air Pollution
h_indoor<- ggplot(high_poll, aes(x=country, y = avg_indoor, cex=5)) + 
  geom_point() +
  ggtitle("Indoor Air Pollution Deaths in High Population Countries")+
  xlab("Country")+
  ylab("Average Deaths")
ggplotly(h_indoor)

#Low Indoor Air Pollution
l_indoor<- ggplot(low_poll, aes(x=country, y = avg_indoor, cex=5)) + 
  geom_point() +
  ggtitle("Indoor Air Pollution Deaths in Low Population Countries")+
  xlab("Country")+
  ylab("Average Deaths")
ggplotly(l_indoor)

# High Outdoor Air Pollution
h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone, cex=5)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "violet", aesthetics = "colour") +
  geom_point() +
  labs(title = "Outdoor Air Pollution Deaths in High Population Countries") +
  xlab("Country") + 
  ylab("Average Deaths")
ggplotly(h_outdoor)

#Low Outdoor Air Pollution
l_outdoor <- ggplot(low_poll, aes(x = country, y= avg_outdoor, color = avg_ozone, cex=5)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "violet", aesthetics = "colour") +
  geom_point() +
  labs(title = "Outdoor Air Pollution Deaths in Low Population Countries")+
  xlab("Country")+
  ylab("Average Deaths")
ggplotly(l_outdoor)
```

***

# Pollution Over Time 
> Let's look at the previous two decades and compare the death count

> has there been a change?

This is the first decade 1996-2006
```{r, echo=FALSE, message=FALSE}
# Look at the trend of deaths within the last decade from low and high populated countries
death_96_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 1996, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_96 = mean(total_deaths))

death_96_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 1996, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_96 = mean(total_deaths))

death_01_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2001, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_01 = mean(total_deaths))

death_01_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2001, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_01 = mean(total_deaths))

death_06_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_06 = mean(total_deaths))

death_06_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2006, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_06 = mean(total_deaths))

#cohesive table so it's easier to visualize 
#Low Population Table 2007-2017
all_low_first <- death_96_low %>% 
  mutate(death_01_low) %>% 
  mutate(death_06_low)

#High Population Table 2007-2017
all_high_first <- death_96_high %>% 
  mutate(death_01_high) %>% 
  mutate(death_06_high)

kable(list(all_high_first))
kable(list(all_low_first))
```
***

This is the second decade 2007-2017
```{r, echo=FALSE, message=FALSE}
# Look at the trend of deaths within the last decade from low and high populated countries
death_07_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2007, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_07 = mean(total_deaths))

death_07_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2007, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_07 = mean(total_deaths))

death_12_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2012, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_12 = mean(total_deaths))

death_12_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2012, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_12 = mean(total_deaths))

death_17_high<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(High_Deaths_17 = mean(total_deaths))

death_17_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2017, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_17 = mean(total_deaths))

#cohesive table so it's easier to visualize 
#Low Population Table 2007-2017
all_low_second <- death_07_low %>% 
  mutate(death_12_low) %>% 
  mutate(death_17_low)

#High Population Table 2007-2017
all_high_second <- death_07_high %>% 
  mutate(death_12_high) %>% 
  mutate(death_17_high)

kable(list(all_high_second))
kable(list(all_low_second))
```

***
Let's see if there is variation by continent.
Here are some tables for the first decade (1996-2006) and second decade (2007-2017) grouped by continent.

```{r}
#North America 1996-2006
north_96 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

north_01 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

north_06 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(north_96,north_01,north_06), caption = "North America Deaths 1996-2006")

# North America 2007-2017
north_07 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

north_12 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

north_17 <- na.omit(north_america)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(north_07,north_12,north_17), caption = "North America Deaths 2007-2017")

#South America 1996-2006
south_96 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

south_01 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

south_06 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")

# South America 2007-2017
south_07 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

south_12 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

south_17 <- na.omit(south_america)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(south_07,south_12,south_17), caption = "South America Deaths 2007-2017")


# Africa 1996-2006
africa_96 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

africa_01 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

africa_06 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(africa_96,africa_01,africa_06), caption = "Africa Deaths 1996-2006")

# Africa 2007-2017
africa_07 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

africa_12 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

africa_17 <- na.omit(africa)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")

#Europe 1996-2006
europe_96 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

europe_01 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

europe_06 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(europe_96,europe_01,europe_06), caption = "Europe Deaths 1996-2006")

#Europe 2007-2017
europe_07 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

europe_12 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

europe_17 <- na.omit(europe)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(europe_07,europe_12,europe_17), caption = "Europe Deaths 2007-2017")


#Asia 1996-2006
asia_96 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

asia_01 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

asia_06 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(asia_96,asia_01,asia_06), caption = "Asia Deaths 1996-2006")

#Asia 2007-2017
asia_07 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

asia_12 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

asia_17 <- na.omit(asia)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(asia_07,asia_12,asia_17), caption = "Asia Deaths 2007-2017")


#Oceania 1996-2006
oceania_96 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 1996) %>% 
  summarize(avg_deaths_96 = mean(total_deaths))

oceania_01 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 2001) %>% 
  summarize(avg_deaths_01 = mean(total_deaths))

oceania_06 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 2006) %>% 
  summarize(avg_deaths_06 = mean(total_deaths))

kable(list(oceania_96,oceania_01,oceania_06), caption = "Oceania Deaths 1996-2006")

#Oceania 2007-2017
oceania_07 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 2007) %>% 
  summarize(avg_deaths_07 = mean(total_deaths))

oceania_12 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 2012) %>% 
  summarize(avg_deaths_12 = mean(total_deaths))

oceania_17 <- na.omit(oceania)  %>% 
  group_by(country) %>% 
  filter(year == 2017) %>% 
  summarize(avg_deaths_17 = mean(total_deaths))

kable(list(oceania_07,oceania_12,oceania_17), caption = "Oceania Deaths 2007-2017")
```
***

Let's graph the previous tables!

The first decade 1996-2006.
```{r, echo=FALSE}
#High Population Deaths 1996 
p_high_first <- all_high_first %>%
  group_by(country) %>% 
  ggplot(aes(x= country, y= High_Deaths_96, fill = country)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
  geom_col(position = 'dodge')

interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths")

ggplotly(interact_high_first)

#High Population Deaths 1996 




#Low Population Deaths 1996-2006
p_low_first <- all_low_first %>%
  group_by(country) %>% 
  ggplot(aes(x= Low_Deaths_96, y= Low_Deaths_01, color= Low_Deaths_06)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
  geom_point() +
  facet_wrap(~country)

interact_low_first<- p_low_first + labs(title="Low Populated Countries", x="1996 Deaths", y="2001 Deaths")

ggplotly(interact_low_first)


#Joined all 
#X - year 
# Avg total deaths 
# color by continent 
```

*** 

This shows the second decade 2007-2017.
```{r, echo=FALSE}
#High Population Deaths 2007-2017
p_high_second <- all_high_second %>%
  group_by(country) %>% 
  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
  geom_point() +
  facet_wrap(~country)

interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")

ggplotly(interact_high_second)

#Low Population Deaths 2007-2017
p_low_second <- all_low_second %>%
  group_by(country) %>% 
  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
  geom_point() +
  facet_wrap(~country)

interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")

ggplotly(interact_low_second)

```

***

> By comparing each pollutant type, we can determine which year and country had the highest numbers of deaths 

Indoor Deaths
```{r, echo=FALSE}
#Indoor 1997-2017
indoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("High Populated Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_high)

indoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Low Populated Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_low)

#Pakistan and Malawi were both mainly affected in 1996
indoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Greatest Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_top)
```

***

Outdoor Deaths
```{r, echo= FALSE}
#Outdoor Particulate 1997-2017
outdoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("High Populated Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_high)

outdoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Low Populated Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_low)

#Serbia and Pakistan were the top countries affected. 
#2011 Pakistan had the greatest amount of Outdoor Deaths
outdoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Greatest Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_top)
```

***
Ozone Deaths 
```{r, echo=FALSE}
#Outdoor ozone 1997-2017
ozone_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("High Populated Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_high)

ozone_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Low Populated Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_low)

#Malawi and Pakistan were the top countries affected 
#1997 had the greatest amount of ozone death in Pakistan
ozone_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Greatest Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_top)
```


# Which is worse? 
> outdoor or indoor pollution?

Let's reintroduce a graph we looked at earlier. Instead this time we will combine the pollutant types together.
```{r, echo=FALSE}
#Plot High Population
h_pollutant<- ggplot(high_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
  geom_point() +
  ggtitle("High Population Pollutant Type Death Counts")
ggplotly(h_pollutant)

#Plot Low Population
l_pollutant<- ggplot(low_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
  geom_point() +
  ggtitle("Low Population Pollutant Type Death Counts")
ggplotly(l_pollutant)
```

We cannot conclude which is worse. 

* High Populated Countries: 
  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.

* Low Populated Countries: 
  + it's half-and-half

# Summary 

* Which country has the highest average death count?
  + High Population: Pakistan
  + Low Population: Malawi
* Has the percentage of the affected population decreased or increased over time?
  + Generally it is decreasing for both High and Low populated countries 
* Which pollutant type has the greatest number of deaths?
  + High Population: Indoor Pollution
  + Low Population: Indoor Pollution
* How has the death count changed over the past two decades?
  + 1996-2006: 
    - High Population: Decreases
    - Low Population: Decreases
  + 2007-2017: 
    - High Population: Decreases 
    - Low Population: Decreases
* Which year and country had the highest number of deaths per pollutant type?
  + We looked at years 1996-2017 
  + Indoor: Pakistan and Malawi were mainly affected in 1996
  + Outdoor: Serbia and Pakistan were the top countries. 
    - 2011 was the worst for Pakistan. 
    - 1997 was the worst for Serbia 
    - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
  + Outdoor Ozone: Pakistan and Malawi were the top countries. 
    - 1997 was the worst for Pakistan. 
    - 1998 was the worst for Malawi
    - United States was the second highest amount of deaths among the higher populated countries
    - Pakistan decreased and then slightly increased
* Which is worse - outdoor or indoor pollution?
  + Inconclusive

# Sources
* [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 ]

* [https://www.worldometers.info/population/world/]

* [https://www.who.int/health-topics/air-pollution#tab=tab_1]

* [https://www.epa.gov/ground-level-ozone-pollution/ground-level-ozone-basics]

* [https://www.health.nsw.gov.au/environment/air/Pages/outdoor-air-pollution.aspx]

* [https://www.mana.md/indoor-air-vs-outdoor-air/]

* [https://www.kaggle.com/datasets/imdevskp/world-population-19602018]

