---
title: "rough outline"
output: slidy_presentation
date: '2022-04-25'
---

# Introduction 
* Air pollution can be detrimental to both our health and the climate
  + Pollutants of major public health concern include carbon monoxide, ozone, nitrogen oxide, and sulfur oxide
  + Outdoor and indoor air pollution cause respiratory and other diseases. 
  + Air pollution kills about 7 million people worldwide every year 

* We want to look at death rates of air pollution world wide

* Why?
  + Hopefully this information can showcase the importance of air pollution and that we should be more mindful about our planet
  
* Overview
  + We will see how different air pollution types affect the population 
  + See if countries have implemented better techniques to reduce air pollution

# Packages Required 

```{r, include=FALSE}
#This will allow us to filter through our data 
library(tidyverse)
library(dplyr)
#This will help us plot figures to showcase our findings
library(ggplot2)
```

# Data Preparation 
The data we will be looking at: 

1. **Deaths due to Air Pollution**
[https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017]

* Author: Akshat Giri
* This data consists the deaths due to air pollution from 1990-2017 around the world
* The dataset was recently updated 2 years ago
* The author wanted to gain more awareness about air pollution and forecast the future 

2. **World Population(pending)**

First let's import the Deaths by Air Pollution data.
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
```

We are going to rename a few of the columns and view the first 6 rows of the data
```{r}
colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")

head(deaths_df)
```

Variables that interest us here include:
* country
* total_deaths
* indoor_deaths
* outdoor_deaths
* ozone_deaths 

Now, we are going to import the World Population data 
```{r}
world_pop <- read.csv("population_total_long.csv")
```

Let's view the first 6 rows of the data
```{r}
head(world_pop)
```
# Exploratory Data Analysis 

# Summary 